`inferCNV` results exploration for `r params$sample_id`

### 2024-11-25 analyses/cell-type-wilms-tumor-06/notebook_template/06_cnv_infercnv_exploration.Rmd

---
title: "`infercnv` results exploration for `r params$sample_id`"
author: "Maud PLASCHKA"
date: "`r Sys.Date()`"
params:
  scpca_project_id: "SCPCP000006"
  sample_id: "SCPCS000179"
  seed: 12345
output: 
  html_document: 
    toc: yes
    toc_float: yes
    code_folding: hide
    highlight: pygments
    df_print: paged
---

@@

This notebook explores using [`infercnv`](https://github.com/broadinstitute/inferCNV/wiki) to estimate tumor and normal cells in `r params$sample_id` from SCPCP000006. 
This sample has a(n) `r subdiagnosis` subdiagnosis.

`infercnv` was run using the `06_inferCNV.R` script with and without a normal reference, from the same patient or from an inter-patient pull of normal cells. We tested the impact of the sub-selection of normal cells using either immune, and/or endothelial cells as healthy reference.
In addition, we are exploring the use of the [HMM based CNV Prediction Methods](https://github.com/broadinstitute/infercnv/wiki/inferCNV-HMM-based-CNV-Prediction-Methods).

`infercnv` currently support two models for HMM-based CNV prediction, what we refer to as the i3 and i6 models. These are set in the `infercnv::run()` as HMM_type='i3' or HMM_type='i6' (i6 is default). Each method operates on the preliminary `infercnv` object which has been processed through the standard inferCNV processing routines, involving subtraction of signal corresponding to "normal (reference)" cells and smoothing operations.

@@

Of note, running `infercnv` with a i6 HMM model is very long, ~2hours of running time per sample, while i3 HMM model only requires ~10min per sample.

@@

For a Seurat object `object`, the function `Do_CNV_heatmap` load the `infercnv` object created with the script `06_infercnv.R` using `reference_value` as a reference and call the function `SCpubr::do_CopyNumberVariantPlot` to plot the mean CNV score in each group defined in `group.by`.

@@

- `infercnv_obj` is the `infercnv` object

@@

- `reference_value` indicates the reference used for `infercnv`, either endothelial, immune, both or pool.

@@

For a `Seurat` object an `infercnv` object created with the script `06_infercnv.R` using `reference_value` as a reference, the function `Do_CNV_score` calculate a CNV score per cell.
The score is calculated based on the [this discussion](https://www.biostars.org/p/9573777/).
The function `Do_CNV_score` returns the `Seurat` object with an additional metadata named `CNV-score_{reference_value}.`

- `reference_value` is the selection of normal cells used for `infercnv`
- `seurat_object` and `infercnv_obj` are the `Seurat`and `infercnv` object created with the script `06_infercnv.R`

@@

#### Visualize seurat clusters and markers genes

@@

#### Wrapper function to explore `infercnv` HMM CNV prediction

The `wrapper_explore_hmm` take as input the `infercnv_obj` generated with `infercnv` HMM CNV predictions.
The wrapper allows the following steps and plots:

@@

For each chromosome, we look at the repartition of the `proportion_cnv_` in cells labeled as immune, endothelial, stroma and fetal nephron.
`proportion_cnv_` is the proportion in number of genes that are part of any cnv/loss/duplication in the given chr.

@@

For each chromosome, we look at the distribution of the `proportion_cnv_` in cells labeled as immune, endothelial, stroma and fetal nephron.
`proportion_cnv_` is the proportion in number of genes that are part of any cnv/loss/duplication in the given chr.

@@

### Heatmap of `infercnv` results

Here we plot the output of `infercnv` as heatmaps of CNV.
We first look at the PNG file generated by the `infercnv` function.
We then used the `infercnv object` to look at mean CNV value across compartments (immune, endothelial, stroma and fetal nephron).

@@

Of note: By default if no reference is provided, `infercnv` take the mean of expression as normal reference. 
The risk is that the main cell population (in our case the fetal nephron compartment) might be mistaken as the normal baseline.

@@

We want to calculate a single CNV score and asesss if/how it can be use to define cells with CNV versus stable/normal cells. 
We defined the score as described in the [biostar discussion](https://www.biostars.org/p/9573777/).

@@

We then explore `infercnv` results generated with immune and endothelial cells as reference, using a [HMM-i3 prediction models](https://github.com/broadinstitute/infercnv/wiki/infercnv-i3-HMM-type).

@@

We then explore `infercnv` results generated with immune and endothelial cells as reference, using a [HMM-i6 model](https://github.com/broadinstitute/infercnv/wiki/infercnv-i6-HMM-type).

@@

We lastly explore `infercnv` results generated with a pull of immune and endothelial cells from different Wilms tumor samples as reference, using a HMM-i3 prediction model.

The selection of normal cells to be spiked-in as a normal reference prior running `infercnv` has been performed in `06b_build-normal-cell_reference.R.
We selected samples that haven't been pre-treated with chemotherapies. Indeed, even if normal cells shouldn't be affected by chemotherapy in terms of CNV, we are not sure how chemotherapies can affect the CNV profile of normal cells.
We thus decided to only take endothelial and immune cells from non-treated samples whose label transfer scores are greater than 0.85 to build the reference of normal cells.

@@

We load the `Seurat` object generated in `06_infercnv.R` and transfer the `infercnv` metadata to the main `Seurat` object of sample `r params$sample_id` saved in `r glue::glue("02b-fetal_kidney_label-transfer_", params$sample_id, ".Rds")`.

@@

- We should run `infercnv` with the most diverse normal reference as possible, including different cell types.
Providing no reference is not a good option, as we think that most of the cells are cancer cells with few CNV. 
In our case, we advise taking at least immune and endothelial cells as normal reference.

- The HMM prediction models help exploring the `infercnv` results. 
In this notebook, we have compared three HMM prediction models:

@@

- Surprisingly, running `infercnv` with endothelial and immune cells from (i) the same patient or (ii) a set of Wilms tumor patients do not seem to affect drastically the results. 
Some false positive CNV might occur in every patient due to the inter-patient variability. 
By comparing the results in conditions (i) and (ii), we should be able to understand which false positive are recurrent and do not take them into account.

>>>>

---
title: "`inferCNV` results exploration for `r params$sample_id`"
author: "Maud PLASCHKA"
date: "`r Sys.Date()`"
params:
  scpca_project_id: "SCPCP000006"
  sample_id: "SCPCS000179"
  seed: 12345
output:
  html_document:
    toc: yes
    toc_float: yes
    code_folding: hide
    highlight: pygments
    df_print: paged
---

@@

This notebook explores using [`inferCNV`](https://github.com/broadinstitute/inferCNV/wiki) to estimate tumor and normal cells in `r params$sample_id` from SCPCP000006.
This sample has a(n) `r subdiagnosis` subdiagnosis.

`inferCNV` was run using the `06_inferCNV.R` script with and without a normal reference, from the same patient or from an inter-patient pull of normal cells. We tested the impact of the sub-selection of normal cells using either immune, and/or endothelial cells as healthy reference.
In addition, we are exploring the use of the [HMM based CNV Prediction Methods](https://github.com/broadinstitute/infercnv/wiki/inferCNV-HMM-based-CNV-Prediction-Methods).

`inferCNV` currently support two models for HMM-based CNV prediction, what we refer to as the i3 and i6 models. These are set in the `infercnv::run()` as HMM_type='i3' or HMM_type='i6' (i6 is default). Each method operates on the preliminary `inferCNV` object which has been processed through the standard `inferCNV` processing routines, involving subtraction of signal corresponding to "normal (reference)" cells and smoothing operations.

@@

Of note, running `inferCNV` with a i6 HMM model is very long, ~2hours of running time per sample, while i3 HMM model only requires ~10min per sample.

@@

For a Seurat object `object`, the function `Do_CNV_heatmap` load the `inferCNV` object created with the script `06_infercnv.R` using `reference_value` as a reference and call the function `SCpubr::do_CopyNumberVariantPlot` to plot the mean CNV score in each group defined in `group.by`.

@@

- `infercnv_obj` is the `inferCNV` object

@@

- `reference_value` indicates the reference used for `inferCNV`, either endothelial, immune, both or pool.

@@

For a `Seurat` object an `inferCNV` object created with the script `06_infercnv.R` using `reference_value` as a reference, the function `Do_CNV_score` calculate a CNV score per cell.
The score is calculated based on the [this discussion](https://www.biostars.org/p/9573777/).
The function `Do_CNV_score` returns the `Seurat` object with an additional metadata named `CNV-score_{reference_value}.`

- `reference_value` is the selection of normal cells used for `inferCNV`
- `seurat_object` and `infercnv_obj` are the `Seurat`and `inferCNV` object created with the script `06_infercnv.R`

@@

#### Visualize Seurat clusters and markers genes

For a Seurat object `object` and a features `features`, the function `visualize_feature` will plot `FeaturePlot` and `ViolinPlot`

@@

#### Wrapper function to explore `inferCNV` HMM CNV prediction

The `wrapper_explore_hmm` take as input the `infercnv_obj` generated with `inferCNV` HMM CNV predictions.
The wrapper allows the following steps and plots:

@@

For each chromosome, we look at the repartition of the `proportion_cnv_` in cells labeled as immune, endothelial, stroma and fetal nephron.
`proportion_cnv_` is the proportion in number of genes that are part of any CNV/loss/duplication in the given chr.

@@

For each chromosome, we look at the distribution of the `proportion_cnv_` in cells labeled as immune, endothelial, stroma and fetal nephron.
`proportion_cnv_` is the proportion in number of genes that are part of any CNV/loss/duplication in the given chr.

@@

### Heatmap of `inferCNV` results

Here we plot the output of `inferCNV` as heatmaps of CNV.
We first look at the PNG file generated by the `inferCNV` function.
We then used the `infercnv object` to look at mean CNV value across compartments (immune, endothelial, stroma and fetal nephron).

@@

Of note: By default if no reference is provided, `inferCNV` take the mean of expression as normal reference.
The risk is that the main cell population (in our case the fetal nephron compartment) might be mistaken as the normal baseline.

@@

We want to calculate a single CNV score and assess if/how it can be used to define cells with CNV versus stable/normal cells.
We defined the score as described in the [discussion](https://www.biostars.org/p/9573777/).

@@

We then explore `inferCNV` results generated with immune and endothelial cells as reference, using a [HMM-i3 prediction models](https://github.com/broadinstitute/infercnv/wiki/infercnv-i3-HMM-type).

@@

We then explore `inferCNV` results generated with immune and endothelial cells as reference, using a [HMM-i6 model](https://github.com/broadinstitute/infercnv/wiki/infercnv-i6-HMM-type).

@@

We lastly explore `inferCNV` results generated with a pull of immune and endothelial cells from different Wilms tumor samples as reference, using a HMM-i3 prediction model.

The selection of normal cells to be spiked-in as a normal reference prior running `inferCNV` has been performed in `06b_build-normal-cell_reference.R.
We selected samples that haven't been pre-treated with chemotherapies. Indeed, even if normal cells shouldn't be affected by chemotherapy in terms of CNV, we are not sure how chemotherapies can affect the CNV profile of normal cells.
We thus decided to only take endothelial and immune cells from non-treated samples whose label transfer scores are greater than 0.85 to build the reference of normal cells.

@@

We load the `Seurat` object generated in `06_infercnv.R` and transfer the `inferCNV` metadata to the main `Seurat` object of sample `r params$sample_id` saved in `r glue::glue("02b-fetal_kidney_label-transfer_", params$sample_id, ".Rds")`.

@@

- We should run `inferCNV` with the most diverse normal reference as possible, including different cell types.
Providing no reference is not a good option, as we think that most of the cells are cancer cells with few CNV.
In our case, we advise taking at least immune and endothelial cells as normal reference.

- The HMM prediction models help exploring the `inferCNV` results.
In this notebook, we have compared three HMM prediction models:

@@

- Surprisingly, running `inferCNV` with endothelial and immune cells from (i) the same patient or (ii) a set of Wilms tumor patients do not seem to affect drastically the results.
Some false positive CNV might occur in every patient due to the inter-patient variability.
By comparing the results in conditions (i) and (ii), we should be able to understand which false positive are recurrent and do not take them into account.
